---
title: "Fish Biodiversity Patterns"
subtitle: "Gulf of Mexico Results"
author: "Hannah L. Owens"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---

```{r setup, include=FALSE}
library(dplyr)
library(sf)
library(terra)
library(latticeExtra)
library(ggplot2)
library(vegan)
library(ggforce)
library(voluModel)
library(tidyterra)
library(metR)
library(indicspecies)
library(stringr)
library(ggpubr)
library(grid)
library(ggthemes)
library(car)

sf::sf_use_s2(FALSE)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Dropbox/DeepFishOccurrences/")

theme_Publication <- function(base_size=10, base_family="sans") {
      (theme_foundation(base_size=base_size, base_family=base_family) + 
         theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5, 
                                         margin = margin(0,0,20,0)),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line.x = element_line(colour="black"),
               axis.line.y = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
       ))
      
}
```

# Readying the data

Read in data and trim occurrences to area of interest--the Gulf of Mexico.

```{r read in data and tidy up}
fishData <- data.frame(read.csv("data/processedFishDataWithDatesAndNames.csv"))
fishHorizonComplete <- fishData %>%
  select(-c("dwc_year", "dwc_month", "dwc_day")) %>% 
  drop_na() %>% 
  filter(dwc_decimalLongitude > -100) %>% 
  filter(-70 > dwc_decimalLongitude) %>% 
  filter(35 > dwc_decimalLatitude)  %>% 
  filter(dwc_decimalLatitude > 16)
fishHorizonComplete <- vect(fishHorizonComplete, 
                            geom = c("dwc_decimalLongitude", 
                                     "dwc_decimalLatitude"))
crs(fishHorizonComplete) <- crs("epsg:4326")
land <- rnaturalearth::ne_countries(scale = "small", 
                                    returnclass = "sv")
land <- project(land, crs("epsg: 4326"))
land <- aggregate(land)
landFish <- is.related(fishHorizonComplete, land, "within")
fishHorizonComplete <- fishHorizonComplete[!landFish,]

gomShapefile <- vect("data/GOM/GOM.shp")
gomShapefile <- project(gomShapefile, crs(fishHorizonComplete))
gomShapefile <- buffer(gomShapefile, 10000)
gomExtent <- ext(gomShapefile)
fishHorizonComplete <- crop(fishHorizonComplete, gomShapefile)

landFish <- is.related(fishHorizonComplete, land, "within")
fishHorizonComplete <- fishHorizonComplete[!landFish,]

gulfFish <- is.related(fishHorizonComplete, gomShapefile, "within")
fishData <- fishHorizonComplete[gulfFish]
pointMap(as.data.frame(geom(fishData)), 
         spName = "All Gulf of Mexico\nFish Occurrences in iDigBio",
         land = land)
```

# Mapping

First, we have to set things up by making a hexagon grid to aggregate the data. This was adapted and updated from <https://strimas.com/post/hexagonal-grids/>.

```{r hex map, warning = FALSE}
# "EPSG:6579 NAD83(2011) / Texas-Centric Albers Equal Area"
fishData <- project(fishData, "epsg:6579")
gomShapefile <- project(gomShapefile, crs(fishData))
land <- project(land, crs(fishData))

# Study area
studyArea <- ext(gomShapefile)
HexPols2 <- st_make_grid(studyArea , 
                         cellsize = 100000, 
                         square = FALSE)
GOM <- vect(HexPols2)
crs(GOM) <- crs(land)
```

## Sampling Density

How many records are there in each hexagonal cell?

```{r sampling density}
# Gridding it
point_density <- extract(x = GOM, y = fishData)
point_density <- table(point_density[,2])
counts <- NULL
for ( i in 1:length(GOM)){
  if (i %in% as.numeric(names(point_density))){
    counts <- c(counts, 
                point_density[names(point_density) == i])
  } else{
    counts <- c(counts, NA)
  }
}

GOM$counts <- counts

my_breaks <- c(1, 10, 50, 100, 500, 1000, 5000, 6000)
my_col <- rgb(0, 0, 255, max = 255, 
              alpha = 0, names = "clear")

#pdf(file = "Figures/GOMDiversity.pdf")
ggplot() + 
  geom_spatvector(data = land, 
                  color = "gray50", 
                  alpha = .5) +
  geom_spatvector(data = GOM, 
                  aes(fill = counts), 
                  color = NA) +
  scale_fill_viridis_b(name = "Records", option = "A", 
                       na.value = my_col, breaks = my_breaks, 
                       limits = c(0, max(my_breaks))) +
  theme_minimal() + 
  coord_sf(crs = st_crs(4326), expand = TRUE,
           xlim = gomExtent[1:2], ylim = gomExtent[3:4], 
           label_graticule = "SW") +
  ggtitle("Gulf of Mexico Sampling Density of\nMarine Bony Fishes, iDigBio") +
  scale_x_longitude(ticks = 5) + scale_y_latitude(ticks = 5)
#dev.off()
```

## Depth Coverage
This map highlights regions of the ocean with differing depth coverage of sampling effort - specifically the percentage of "available depth" in each hexagon covered by records in iDigBio.

```{r depth sampling, warning=FALSE}
# Get bathymetery
bathymetry <- rast("data/ETOPO_2022_v1_60s_N90W180_bed.tif") %>%
  crop(ext(project(GOM, crs("epsg:4326")))) %>% 
  project(crs(GOM), method = "min")
GOM$bathymetry <- extract(x = bathymetry, y = GOM, fun = min)[,2]
GOM$bathymetry[GOM$bathymetry > 0] <- NA

# Get sampling depth
samplingDepth <- NULL
for (i in 1:length(GOM)){
  refs <- is.related(fishData, GOM[i], "within")
  if(any(refs)){
    maxDepth <- max(fishData[refs,"dwc_maximumDepthInMeters"], 
                    na.rm = TRUE)
  } else {
    maxDepth <- NA
  }
  samplingDepth <- c(samplingDepth, maxDepth)
}

GOM$maxSample <- -samplingDepth
GOM$samplingRatio <- GOM$maxSample/GOM$bathymetry
GOM$samplingRatio[GOM$samplingRatio > 1] <- 1

# Put it all together
#pdf("Figures/GOMSamplingRatio.pdf")
ggplot() +
  geom_spatvector(data = land, 
                  color = "gray50", 
                  alpha = .5) +
  geom_spatvector(data = GOM, 
                  aes(fill = samplingRatio), 
                  color = NA) +
  scale_fill_viridis_b(name = "Ratio", option = "A", 
                       na.value = my_col, 
                       breaks = seq(0,10, by =1)/10) +
  theme_minimal() + 
  coord_sf(crs = st_crs(4326), expand = TRUE,
           xlim = gomExtent[1:2], ylim = gomExtent[3:4], 
           label_graticule = "SW") +
  ggtitle("Gulf of Mexico Sampling vs Available Depth Ratio\nof Marine Bony Fishes, iDigBio") +
  scale_x_longitude(ticks = 5) + 
  scale_y_latitude(ticks = 5)
#dev.off()
```

# Diversity Pattern Analyses

First, we do some additional filtering to fit the data for our use, including removing records with excessive (i.e. 100m) uncertainty in depth.

```{r uncertainty filtering, echo = FALSE}
# Remove records with excessive uncertainty
print(paste0("Before removing records with a depth uncertainty of 100m or more, there are ", nrow(fishData), " records."))
fishData$uncertainty <- fishData$dwc_maximumDepthInMeters - fishData$dwc_minimumDepthInMeters
fishData <- fishData[fishData$uncertainty < 100,]
print(paste0("After removing records with a depth uncertainty of 100m or more, there are ", nrow(fishData), " records."))
fishData <- fishData[str_detect(fishData$dwc_scientificName, " sp\\.", negate = TRUE),]
fishData <- fishData[str_detect(fishData$dwc_scientificName, "species", negate = TRUE),]
print(paste0("After removing records that were not identified to species, there are ", nrow(fishData), " records."))
```

Next, we assign each record to equal-width depth bins under two strategies: 1) No empty bins permitted, 2) At least 10 records per bin.

```{r determining bin number}
# Zero empty bins
breakNumber <- 60 
test <- TRUE
while (test){
    breakNumber <- breakNumber - 1
    tmp <- hist(fishData$depth,  
                breaks = breakNumber, 
                plot = F)
    test <- 0 %in% tmp$counts
}
zeroEmpty <- length(tmp$breaks)

# Minimum ten per bin
breakNumber <- 60 
test <- TRUE
while (test){
    breakNumber <- breakNumber - 1
    tmp <- hist(fishData$depth,  
                breaks = breakNumber, 
                plot = F, )
    test <- 10 >= min(tmp$counts)# Checks to see if any bins are less than ten
}
tenPerBin <- length(tmp$breaks)

paste0("Number of bins if no bin can be empty: ", 
       zeroEmpty)
paste0("Number of bins if each bin must have at least 10 observations: ", 
       tenPerBin)

rm(tmp, breakNumber, test)
```

## Calculate diversity statistics

Next, we define a function to calculate diversity statistics we need for downstream analysis.

```{r define function to calculate bin stats, fig.show='hide', message=FALSE, warning=FALSE}
fishData <- data.frame(fishData)

diversityStats <- function(x, binNumber){
  # Get breaks based on equal width bins
  breaks <- seq(min(x$depth), 
              max(x$depth), 
              (max(x$depth) - min(x$depth))/(binNumber))
  
  # Species richness per bin
  depthDiversityCount <- NULL
  effort <- NULL
  for(i in 2:length(breaks)){
    temp <- x %>% filter(depth >= breaks[i-1], 
                         breaks[i] >= depth)
    depthDiversityCount <- c(depthDiversityCount,
                             length(unique(temp$dwc_scientificName)))
    effort <- c(effort, nrow(temp))
  }
  x$depthBin <- cut(x$depth, breaks=breaks, 
                    labels = paste0(round(breaks[-(binNumber+1)], 
                                          digits = 0), " to ", 
                                    round(breaks[-1], 
                                          digits = 0)))
  
  # Compile data -- Spp are columns, sites are rows, counts are cells
  depthComm <- data.frame(table(data.frame(x[c("depthBin", 
                                               "dwc_scientificName")])))
  depthComm <- as.data.frame.matrix(xtabs(Freq ~ depthBin + dwc_scientificName, 
                                          depthComm))
  diversityResults <- data.frame(row.names = row.names(depthComm))
  diversityResults$bin <- seq(1, nrow(depthComm))
  
  # Calculate dissimilarity among depth bands
  depthComm.pa <- decostand (depthComm, method = 'pa')
  dis <- vegdist(depthComm, "chao") # Tries to account for unseen species pairs
  
  # Calculate clustering
  clust.res<-hclust(dis, "ward.D2")
  clus_cut <- cutree (clust.res, k = 2)
  phi <- multipatt (depthComm.pa, 
                    cluster = clus_cut, 
                    fun = 'r.g')
  phi$sign <- phi$sign[complete.cases(phi$sign),]
  re <- phi$sign[phi$sign$p.value<=0.01,]
  
  # Ordinate and prepare for plotting
  mds <- cmdscale(dis, eig = TRUE)
  mds$species <- wascores(mds$points, 
                          depthComm, 
                          expand = TRUE)
  pl <- ordiplot(mds, disp = "sites", 
                 type = "none")
  if (binNumber > 2){
    clusters <- kmeans(dis, centers = 2)
    depthCluster <- as.factor(clusters$cluster)
  } else {
    clusters <- kmeans(dis, centers = 1)
    depthCluster <- as.factor(clusters$cluster)
  }

  sites.long <- BiodiversityR::sites.long(pl, 
                                          env.data = depthCluster)
  sites.long[, 2:3] <- sapply(sites.long[, 2:3], 
                              as.numeric)
  env.data <- as.factor(depthCluster)
  axis.long <- BiodiversityR::axis.long(mds, 
                                        choices=c(1, 2), 
                                        cmdscale.model = T)
  centroids.long <- BiodiversityR::centroids.long(y = sites.long, 
                                                  grouping=env.data, 
                                                  FUN = "median")
  
  # Put together all the useful outputs
  out <- list("breaks" = breaks, 
              "depthDiversityCount" = depthDiversityCount, 
              "effort" = effort, "data" = x, 
              "binNumber" = binNumber, "depthComm" = depthComm,
              "dis" = dis, "clust.res" = clust.res, 
              "clusters" = clusters, "spClusters" = re,
              "env.data" = env.data, "sites.long" = sites.long, 
              "axis.long" = axis.long, 
              "centroids.long" = centroids.long)
  return(out)
}

# Do the calculations
zeroEmpty <- diversityStats(x = fishData, binNumber = zeroEmpty)
tenPerBin <- diversityStats(x = fishData, binNumber = tenPerBin)
```

## Species richness vs depth

For each binning strategy, we initially plotted diversity at depth. We include lines at 200m and 1000m, which are commonly-used boundaries for epipelagic and mesopelagic zone boundaries, respectively.

```{r depth vs diversity}
# Plot diversity by depth
plot(cbind(zeroEmpty$breaks[-(length(zeroEmpty$breaks))], 
           zeroEmpty$depthDiversityCount), 
     type = "l", xlab = "Depth (m)", ylab = "Species Richness",
     main = paste0("Species Richness vs Depth\nNo Empty Bins (", 
                   zeroEmpty$binNumber, " Bins)"))
abline(v = 200, col = "darkred", lty = "dashed")
abline(v = 1000, col = "darkred", lty = "dashed")

plot(cbind(tenPerBin$breaks[-(length(tenPerBin$breaks))], 
           tenPerBin$depthDiversityCount), 
     type = "l", xlab = "Depth (m)", ylab = "Species Richness",
     main = paste0("Species Richness vs Depth\nAt Least 10 per Bin (", 
                   tenPerBin$binNumber, " Bins)"))
abline(v = 200, col = "darkred", lty = "dashed")
abline(v = 1000, col = "darkred", lty = "dashed")
```


## Null Model for Diversity

How does the observed distribution of diversity compare to a null distribution? That is, what if any observation was equally likely to be found in each depth bin--do we see the same distribution of diversity?

```{r null diversity test}
calcEffortAndAlpha <-function(rawData){
  effortAndAlpha <- data.frame(table(rawData$data$depthBin))
  for(i in 1:nrow(effortAndAlpha)){
    binCommunity <- rawData$data[rawData$data$depthBin == effortAndAlpha$Var1[i],]
    binCommunity <- binCommunity[complete.cases(binCommunity$depthBin),]
    effortAndAlpha$alpha[i] <- length(unique(binCommunity$dwc_scientificName))
  }
  colnames(effortAndAlpha) <- c("Depth", "Count", "Species_Richness")
  return(effortAndAlpha)
}

# Simulate data
fishDatSim <- tenPerBin
fishDatSim$data <- fishDatSim$data[!is.na(fishDatSim$data$depthBin),]
fishDatSim <- lapply(1:100, FUN = function(x) {
    fishDatSim$data$depthBin <- sample(fishDatSim$data$depthBin)
    return(fishDatSim)
  })
# Calculate species richness
allDat <- lapply(fishDatSim, 
                 FUN = function(x) calcEffortAndAlpha(x))

effortAndAlpha <-calcEffortAndAlpha(tenPerBin)

# Plot sims and obs
P1 <- ggplot()
medProp <- c()
for(dat in allDat){
  P1 <- P1 +
    geom_line(data = dat, aes(x = Depth, y = Species_Richness, group = 1), 
              alpha=0.01)
  medProp <- c(medProp, median(dat$prop))
}

finalPlot <- P1 +
  geom_line(data = effortAndAlpha, 
            aes(x = Depth, y = Species_Richness, group = 1), 
            col = "red") +
  theme_Publication() + 
  labs( x = "Depth (m)", y = "Species Richness") +
  theme(axis.text.x = element_text(angle = 45, hjust=.5, 
                                   vjust = .5)) +
  ggtitle("Depth versus Diversity")

finalPlot

ggsave("Figures/DiversityObsVsSimGOM.png",finalPlot,height = 8.7, width = 11.4, units = "cm")
```
## Possible correlates with species richness

### Number of records and area covered

Of course, there are plenty of non-biological reasons diversity might decrease with depth. Two of our hypothesized correlations were decreased sampling effort and amount of available area as depth increases.

```{r Sampling effort versus species richness}
effortAndAlpha$availableArea <- NA
gomBath <- mask(x = bathymetry, mask= gomShapefile)
cellBath <- cellSize(gomBath, unit = "km")
for (i in 1:nrow(effortAndAlpha)){
  depths <- as.numeric(unlist(strsplit(as.character(effortAndAlpha$Depth[i]), 
                                       " to ")))
  effortAndAlpha$availableArea[i] <- unlist(global(((gomBath < -(depths[1])) * cellBath), 
                                                   sum, na.rm = T))
}
```

### Correlation between clines and diversity

It was suggested by Katherine Richardson, Owens's colleague at the University of Copenhagen, that the bump in the diversity decay curve could be the result of thermoclines and/or pycnoclines. We used temperature and salinity data from the World Ocean Atlas 2018 dataset to investigate this idea, extracting mean temperature and salinity for each depth bin and comparing it to diversity. We also examined apparent oxygen utilization and nitrogen concentration as potential correlates, as these variables are also available via the WOA 2018 dataset and may be correlated with diversity.

```{r environmental variable extraction}
# Define extraction function
envDatExtract <- function(envRaster, mShp, breaks){
  # Get all the points
  dummyOccs <- data.frame(0, 0, 0)
  colnames(dummyOccs) <- c("longitude", "latitude", "depth")
  grid <- mSampling3D(occs = dummyOccs, 
                      envBrick = envRaster, 
                      mShp = mShp, verbose = F)
  allData <- cbind(grid, 
                 xyzSample(grid, envBrick = envRaster, verbose = F))
  colnames(allData) <- c("longitude", "latitude", "depth", "Variable")
  
  # Assign points to bins and calculuate summary stats
  binSampledData <- NULL
  for(i in 2:length(breaks)){
    temp <- allData %>% filter(depth >= breaks[i-1], breaks[i] >= depth)
    binSampledData <- c(binSampledData, mean(temp[,4]))
  }
  return(binSampledData)
}

effortAndAlpha$"Available Area" <-effortAndAlpha$availableArea
effortAndAlpha <- select(effortAndAlpha, -c("availableArea"))

effortAndAlpha$Temperature <- rast("~/Dropbox/MARDIGRA/data/EnvironmentalData/ProcessedEnvtData/temperature.tif") %>%
  project(crs(gomShapefile)) %>% 
  crop(gomShapefile) %>% 
  mask(mask = gomShapefile) %>% 
  envDatExtract(mShp = gomShapefile, breaks = tenPerBin$breaks)
effortAndAlpha$Salinity <- rast("~/Dropbox/MARDIGRA/data/EnvironmentalData/ProcessedEnvtData/salinity.tif")  %>% 
  project(crs(gomShapefile)) %>% 
  crop(gomShapefile) %>% 
  mask(mask = gomShapefile) %>% 
  envDatExtract(mShp = gomShapefile, breaks = tenPerBin$breaks)
effortAndAlpha$AOU <- rast("~/Dropbox/MARDIGRA/data/EnvironmentalData/ProcessedEnvtData/AOU.tif") %>%
  project(crs(gomShapefile)) %>% 
  crop(gomShapefile) %>% 
  mask(mask = gomShapefile) %>% 
  envDatExtract(mShp = gomShapefile, breaks = tenPerBin$breaks)
effortAndAlpha$Nitrate <- rast("~/Dropbox/MARDIGRA/data/EnvironmentalData/ProcessedEnvtData/nitrate.tif") %>%
  project(crs(gomShapefile)) %>% 
  crop(gomShapefile) %>% 
  mask(mask = gomShapefile) %>% 
  envDatExtract(mShp = gomShapefile, breaks = tenPerBin$breaks)
```

### Plotting

Finally, we plot alpha diversity and its hypothesized correlates to identify emergent trends.

```{r, fig.height = 10}
effortAndAlphaLong <- reshape2::melt(effortAndAlpha[,c("Depth", "Species_Richness", 
                                                        "Count", "Available Area",
                                                        "Temperature", "Salinity", 
                                                        "AOU", "Nitrate")], 
                                      id.vars = "Depth")

P1 <- ggplot(effortAndAlphaLong, aes(Depth, value, 
                                      colour = variable, group = 1)) +
  geom_line() +
  theme_Publication() + 
  labs( x = "Depth (m)") +
  theme(axis.text.x = element_text(angle = 45, hjust=.5, vjust = .5)) +
  scale_color_brewer(palette = "Dark2") +
  facet_wrap(variable~., ncol = 1, scales = "free_y") + 
  theme(legend.position="none", axis.title.y = element_blank())
P1
ggsave("Figures/GOMClines.png",P1,height = 22.5, width = 8.7, units = "cm")
```

# Summary statistics

Finally, here are some sampling stats.

```{r a few summary stats, echo=FALSE}
print(paste0("There are ", length(unique(fishData$dwc_scientificName)), 
             " species in the GOM dataset."))

print(paste0("There are ", nrow(GOM), 
             " cells in the GOM dataset."))
print(paste0(nrow(GOM[GOM$counts > 9]), 
             " cells in the GOM dataset have at least 10 records."))
print(paste0(nrow(GOM[GOM$counts > 99]), 
             " cells in the GOM dataset have at least 100 records."))

print(paste0(nrow(GOM[GOM$bathymetry <= -200]), 
             " cells in the GOM dataset have bathymetry at or exceeding 200m."))
GOMFiltered <- GOM %>% filter(bathymetry <= -200) %>% 
    filter(maxSample <= -200) 
print(paste0(nrow(GOMFiltered), 
             " GOM cells (", round((nrow(GOMFiltered)/nrow(GOM))*100, digits = 0), "%) have bathymetry at or exceeding 200m AND have at least one record from at or below 200m."))
```

# Appendix: Additional analyses 

The following analyses were also initially performed, but were not conclusive enough to include in the final manuscript. We include them here should they prove useful to other researchers with similar questions.

## Cluster analysis

One of the first questions we were interested in addressing with this dataset was whether there was clear community structure with depth. We calculated how similar the species assemblages in each bin were by standardizing the data to a presence/absence matrix, then calculating Chao similarity. The results are plotted as a dendrogram, where shorter branches between tips indicate more similarity between bins. 

```{r dendrogram plots}
# How do they cluster?
plot(zeroEmpty$clust.res, 
     main = paste0("Bin Species Composition Similarity\nZero Empty Bins (",
                   zeroEmpty$binNumber, " Bins)"))
rect.hclust (zeroEmpty$clust.res, k = 2)
rect.hclust (zeroEmpty$clust.res, k = 3, border = "blue")

plot(tenPerBin$clust.res, 
     main = paste0("Bin Species Composition Similarity\nAt Least Ten per Bin (",
                   tenPerBin$binNumber, " Bins)"))
rect.hclust (tenPerBin$clust.res, k = 2)
rect.hclust (tenPerBin$clust.res, k = 3, border = "blue")
```


Next we take the pairwise similarity scores of the bins, and perform multidimensional scaling so that the distances between points are approximately equal to their dissimilarity.

```{r ordination plots}
# Ordinate and plot
BioR.theme <- theme(
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid = element_blank(),
        axis.line = element_line("gray25"),
        text = element_text(size = 12),
        axis.text = element_text(size = 10, 
                                 colour = "gray25"),
        axis.title = element_text(size = 14, 
                                  colour = "gray25"),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 14),
        legend.key = element_blank())

plotDiversityZeroEmpty <- ggplot() + 
    geom_vline(xintercept = c(0), color = "grey70", 
               linetype = 2) +
    geom_hline(yintercept = c(0), color = "grey70", 
               linetype = 2) +  
    xlab(zeroEmpty$axis.long[1, "label"]) +
    ylab(zeroEmpty$axis.long[2, "label"]) +  
    scale_x_continuous(sec.axis = dup_axis(labels=NULL, 
                                           name=NULL)) +
    scale_y_continuous(sec.axis = dup_axis(labels=NULL, 
                                           name=NULL)) +
    geom_mark_hull(data=zeroEmpty$sites.long, 
                   aes(x=axis1, y=axis2, 
                       colour=zeroEmpty$env.data, 
                       fill=after_scale(alpha(colour, 0.2))), 
                   concavity=0.1, linewidth=0.2, show.legend=FALSE) +
    geom_segment(data=zeroEmpty$centroids.long, 
                 aes(x=axis1c, y=axis2c, xend=axis1, 
                     yend=axis2, colour=zeroEmpty$env.data), 
                 linewidth=1, show.legend=FALSE) +
    geom_point(data=zeroEmpty$sites.long, 
               aes(x=axis1, y=axis2, colour=zeroEmpty$env.data, 
                   shape=zeroEmpty$env.data), 
               size=5) +
    ggtitle(paste0("Similarity Among Depth Bins,\nZero Empty Bins (", 
                   zeroEmpty$binNumber, " Bins)")) +
    BioR.theme +
    scale_colour_brewer(type = "qual", palette = "Dark2") +
    coord_fixed(ratio=1)

plotDiversityTenPerBin <- ggplot() + 
    geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
    geom_hline(yintercept = c(0), color = "grey70", linetype = 2) +  
    xlab(tenPerBin$axis.long[1, "label"]) +
    ylab(tenPerBin$axis.long[2, "label"]) +  
    scale_x_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +
    scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +
    geom_mark_hull(data=tenPerBin$sites.long, 
                   aes(x=axis1, y=axis2, colour=tenPerBin$env.data, 
                       fill=after_scale(alpha(colour, 0.2))), 
                   concavity=0.1, linewidth=0.2, show.legend=FALSE) +
    geom_segment(data=tenPerBin$centroids.long, 
                 aes(x=axis1c, y=axis2c, xend=axis1, yend=axis2, 
                     colour=tenPerBin$env.data), 
                 linewidth    =1, show.legend=FALSE) +
    geom_point(data=tenPerBin$sites.long, 
               aes(x=axis1, y=axis2, colour=tenPerBin$env.data, 
                   shape=tenPerBin$env.data), 
               size=5) +
    ggtitle(paste0("Similarity Among Depth Bins,\nAt Least Ten per Bin (", 
                   tenPerBin$binNumber, " Bins)")) +
    BioR.theme +
    scale_colour_brewer(type = "qual", palette = "Dark2") +
    coord_fixed(ratio=1)

plotDiversityZeroEmpty
plotDiversityTenPerBin
```

Both bin schemes show two distinct clusters. Which bins are in these groups?!

```{r bin group identities,  echo = FALSE}
print("Cluster identities for each bin, no empty bins are permitted:")
t(t(zeroEmpty$clusters$cluster))

print("Cluster identities for each bin, minimum of ten records per bin:")
t(t(tenPerBin$clusters$cluster))
```

## Identity test

Finally, are there any species that clearly and signifcantly associate with one or the other of the depth clusters we have identified? To to this, we use `multipatt()`, an implementation of metrics developed in De Cáceres and Legendre (2009).

```{r multipatt() results}
tenPerBin$spClusters[tenPerBin$spClusters$p.value < 0.01,]
```
