---
title: "Raw Data Cleaning"
author: "Hannah L. Owens"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Dropbox/DeepFishOccurrences/")
```

# Processing inputs

First, I read in the data and do some preliminary name cleaning and assessment.

```{r read in data and do preliminary name cleaning}
library(stringr)
library(dplyr)
library(terra)

# Using raw search results; iDigBio processed data does not include latitude or longitude
fishData <- read.csv("data/866ac026-5731-4ca2-bb51-7b154d7cde25/occurrence_raw.csv")

# There are 187 columns in the raw dataset. I select the relevant ones for our purposes.
fishData <- fishData[, c("dwc.family", "dwc.genus", "dwc.scientificName",
                         "dwc.verbatimDepth", "dwc.maximumDepthInMeters", 
                         "dwc.minimumDepthInMeters", "dwc.decimalLongitude", 
                         "dwc.decimalLatitude","dwc.year","dwc.month","dwc.day")]

# Fix blanks
fishData[fishData == ""] <- NA

# Remove records with no depth information whatsoever
fishData <- fishData %>%
  filter_at(vars(dwc.maximumDepthInMeters, dwc.minimumDepthInMeters, dwc.verbatimDepth), any_vars(!is.na(.)))

# Cleaning names
fishData$dwc.scientificName <- gsub(r"{\s*\([^\)]+\)}","",
                                    as.character(fishData$dwc.scientificName))
fishData$dwc.scientificName <- gsub("cf. ","",
                                    as.character(fishData$dwc.scientificName))
fishData$dwc.scientificName <- gsub("c. ","",
                                    as.character(fishData$dwc.scientificName))
fishData$dwc.scientificName <- word(fishData$dwc.scientificName, 1,2, sep=" ")
fishData$dwc.genus <- str_to_title(word(fishData$dwc.genus, 1))

fishData$dwc.family <- toupper(word(fishData$dwc.family, 1))
fishData$dwc.family[fishData$dwc.family == "TRIGILIDAE"] <- "TRIGLIDAE"
fishData$dwc.family[fishData$dwc.family == "ZENIONIDAE"] <- "ZENIONTIDAE"
fishData$dwc.family[fishData$dwc.family == "CHAUNAEIDAE"] <- "CHAUNACIDAE"
fishData$dwc.family[fishData$dwc.family == "ECHENEIDIDAE"] <- "ECHENEIDAE"
fishData$dwc.family[fishData$dwc.family == "ELEOTRIDIDAE"] <- "ELEOTRIDAE"
fishData$dwc.family[fishData$dwc.family == "ENGRAULIDIDAE"] <- "ENGRAULIDAE"
fishData$dwc.family[fishData$dwc.family == "EPHIPPIDIDAE"] <- "EPHIPPIDAE"
fishData$dwc.family[fishData$dwc.family == "GONOSTOMIDAE"] <- "GONOSTOMATIDAE"
fishData$dwc.family[fishData$dwc.family == "MACRORHAMPHOSIDAE"] <- "MACRORAMPHOSIDAE"
fishData$dwc.family[fishData$dwc.family == "MACROURIDAEX"] <- "MACROURIDAE"
fishData$dwc.family[fishData$dwc.family == "MONOCENTRIDIDAE"] <- "MONOCENTRIDAE"
fishData$dwc.family[fishData$dwc.family == "OPHICHTHYIDAE"] <- "OPHICHTHIDAE"
fishData$dwc.family <- str_to_title(fishData$dwc.family)

# Fills in missing genus if scientific name is present
for (i in seq(1, nrow(fishData))){
  species <- fishData$dwc.scientificName[i]
  if(!is.na(species)){
    if(str_detect(pattern = "Leptocephali", species)) fishData$dwc.scientificName[i] <- NA
  }
  genus <- fishData$dwc.genus[i]
  if(!is.na(genus) && (genus == "Gen.")) genus <- NA
  if(is.na(genus)){
      sciName <- fishData$dwc.scientificName[i]
      if(!is.na(sciName)){
        genus <- word(sciName, 1)
        if(!str_detect(string = genus, "idae")){
          fishData$dwc.genus[i] <- genus
        } else if(str_detect(string = genus, "idae")){
          if(is.na(fishData$dwc.family[i])){
            fishData$dwc.family[i] <- genus
            fishData$dwc.genus[i] <- NA
          }
        }
      }
  }
}

fishData <- fishData %>%
  filter_at(vars(dwc.genus, dwc.scientificName), all_vars(!is.na(.)))

fishData <- arrange(fishData, dwc.family, 
                    dwc.genus, dwc.scientificName)

paste0("Number of families: ", 
       length(unique(fishData$dwc.family)))
paste0("Number of genera: ", 
       length(unique(fishData$dwc.genus)))
paste0("Number of species: ",
       length(unique(fishData$dwc.scientificName)))

paste0("Number of entries with no family: ", 
       sum(is.na(fishData$dwc.family)))
paste0("Number of entries with no genus: ", 
       sum(is.na(fishData$dwc.genus)))
paste0("Number of entries with no scientific name: ", 
       sum(is.na(fishData$dwc.scientificName)))
paste0("Number of entries with incomplete max/min depth data (pre-cleaning): ", 
       sum(!complete.cases(fishData[,c("dwc.maximumDepthInMeters", 
                                       "dwc.minimumDepthInMeters")])))
```
Some of the missing name data may be fixable, but that is best done by hand later. For example, in some cases a family is missing, but the scientific name is there, so it can be looked up.

Next, I read in the data and clean the maximum and minimum depth columns, getting rid of units and converting depths  in feet or fathoms to meters. It is assumed that measurements without units are in meters.

```{r process max and min depth values, message=FALSE, warning=FALSE}
# Function to convert units
convertUnits <- function(input, minMax = "both") {
  # Error checking
  if (is.na(input)) return(NA)
  if (!str_detect(input, regex("\\d"))) return(NA)
  if (str_detect(input, "Inf")) return(NA)
  if (str_length(gsub("\\D", "", input)) < 1) return(NA)
  
  # Process input
  input <- tolower(input)
  input <- gsub(x = input, " ", "")
  input <- gsub(x = input, "to", "-")
  input <- gsub(x = input, "ca.", "")
  units <- str_extract(input, "[cfkm]\\w*+")
  input <- gsub(input, pattern = "\\s*\\D+$", replacement = "")
  if(str_detect(string = input, pattern = "-", )){
    input <- unlist(strsplit(input, "-"))
    input <- input[!input == ""]
    input <- suppressWarnings(as.numeric(input))
    if(all(is.na(input))) return(NA)
    input <- input[!is.na(input)]
    if(minMax == "min") input <- min(input)
    if(minMax == "max") input <- max(input)
  }
  
  if(is.na(units)) return(suppressWarnings(as.numeric(input)))
  
  # Conversion
  if (any(str_detect(string = units, pattern = c("feet", "ft", "'")))){
    return(suppressWarnings(as.numeric(gsub("\\D", "", input)) * 0.3048))
  }
  if (any(str_detect(string = units, pattern = c("fm", "fms", "fathoms", "f")))){
    return(suppressWarnings(as.numeric(gsub("\\D", "", input)) * 1.8288))
  }
  if (any(str_detect(string = units, pattern = c("cm")))){
    return(suppressWarnings(as.numeric(gsub("\\D", "", input))/100))
  }
  if (any(str_detect(string = units, pattern = c("m", "meters")))){
    return(suppressWarnings(as.numeric(gsub("\\D", "", input))))
  }
}

# Loop through the rows
for (i in 1:nrow(fishData)) {
  max <- convertUnits(fishData$dwc.maximumDepthInMeters[i], minMax = "max")
  min <- convertUnits(fishData$dwc.minimumDepthInMeters[i], minMax = "min")
  range <- c(min, max)
  
  # Check for max/min inversion and assign to rightful place in the table
  if (!all(is.na(range))){
    fishData$dwc.maximumDepthInMeters[i] <- max(c(max, min), na.rm = TRUE)
    fishData$dwc.minimumDepthInMeters[i] <- min(c(max, min), na.rm = TRUE)
  }
}

paste0("Number of entries with incomplete max/min depth data (post-cleaning): ", 
       sum(!complete.cases(fishData[,c("dwc.maximumDepthInMeters", 
                                       "dwc.minimumDepthInMeters")])))
```

Next, for species without maximum and/or minimum depths, we try to extract additional information from the verbatim depth data. Assumptions again include assuming depths with no unit are in meters, and that if both maximum and minimum depths already existed, they are more accurate than the verbatim depth.

``` {r process verbatim depths}
fishData[fishData == "Inf"] <- NA
fishData[fishData == "-Inf"] <- NA

# Mining verbatim depth for precious missing information
for (i in 1:nrow(fishData)){
  max <- fishData$dwc.maximumDepthInMeters[i]
  min <- fishData$dwc.minimumDepthInMeters[i]
  if(any(is.na(c(min,max)))){
    input <- fishData$dwc.verbatimDepth[i]
    output <- convertUnits(input, minMax = "both")
    range <- c(min, max, output)
    if(!all(is.na(range))){
      fishData$dwc.maximumDepthInMeters[i] <- max(range, na.rm=TRUE)
      fishData$dwc.minimumDepthInMeters[i] <- min(range, na.rm=TRUE)
    }
  }
}

paste0("Number of entries with incomplete max/min depth data (post-mining): ", 
       sum(!complete.cases(fishData[,c("dwc.maximumDepthInMeters", 
                                       "dwc.minimumDepthInMeters")])))
```
Ok, great, we now have more complete depth range records than we started with.

# Filtration

The next step is to remove all the occurrences that are outside of the ocean or have recorded depths that are deeper than 99% of the depth of a given pixel. We chose 99% to remove occurrences with depths that exceed the known depth at a given pixel, but also to account for potential noise in the bathymetry data.

```{r filter records without coordinates or inappropriate depths}
# Get depth
bathymetry <- rast("data/ETOPO_2022_v1_60s_N90W180_bed.tif")
names(bathymetry) <- "bathymetry"
bathymetry[bathymetry > 0] <- NA
bathymetry <- abs(bathymetry)

# Filter out points without coordinates
fishData <- fishData %>% filter(!is.na(dwc.maximumDepthInMeters), 
                                !is.na(dwc.decimalLongitude), 
                                !is.na(dwc.decimalLatitude))

paste0("Number of entries with complete max depth data and occurrence coordinates: ", 
       nrow(fishData))

# Filter out points that are deeper than bathymetry
fishData$mapDepth <- round(terra::extract(bathymetry,
                                             as.matrix(fishData[,c("dwc.decimalLongitude", 
                                                                   "dwc.decimalLatitude")]),
                                             method = "bilinear"),0)
  
# Get rid of depths that are deeper or at the same depth as bathymetry
fishData <- fishData[(as.numeric(fishData$dwc.maximumDepthInMeters)/fishData$mapDepth) < 0.999,]

paste0("Number of entries with maximum depths that do not exceed actual depth at occurrence: ", 
       nrow(fishData))
```

Finally, I calculate a column with a single depth measurement--the median between minimum and maximum depth. In cases where only the minimum or maximum are given, this number is used. I then save the .csv file so it is not necessary to repeat the analysis.

```{r calculate median depth column, dev='png'}
fishData$depth <- (as.numeric(fishData$dwc.maximumDepthInMeters) + 
                     as.numeric(fishData$dwc.minimumDepthInMeters))/2
for(i in 1:nrow(fishData)){
  if(is.na(fishData$depth[i])){
    if(!is.na(fishData$dwc.minimumDepthInMeters[i])){
      fishData$depth[i] <- as.numeric(fishData$dwc.minimumDepthInMeters[i])
    } else if(!is.na(fishData$dwc.maximumDepthInMeters[i])){
      fishData$depth[i] <- as.numeric(fishData$dwc.maximumDepthInMeters[i])
    }
  }
}
fishData$dwc.maximumDepthInMeters <- round(as.numeric(fishData$dwc.maximumDepthInMeters))
fishData$dwc.minimumDepthInMeters <- round(as.numeric(fishData$dwc.minimumDepthInMeters))
fishData$depth <- round(fishData$depth)
summary(fishData$depth)

# Get rid of zero depths and NAs
fishData <- fishData[fishData$depth > 0,]
fishData <- fishData[!is.na(fishData$depth),]

hist(fishData$depth, main = "Occurrences by Depth")

colnames(fishData) <- gsub("\\.", "_", colnames(fishData))

missingFamilies <- fishData[is.na(fishData$dwc_family),]
missingFamilies <- do.call(data.frame,missingFamilies)
write.csv(missingFamilies, "data/missingFamilies.csv", row.names = F)

fishData <- do.call(data.frame,fishData)
write.csv(fishData, 
          "data/processedFishDataWithDates.csv", 
          row.names = F)

paste0("There are ", nrow(fishData), " entries in the dataset.")
```

# Manual Taxonomic Review

As mentioned at the beginning of this analysis, several records had no family provided. These records were reviewed by R. Singer, who also flagged several instances of illegitimate species names. In this final step, I mapped his feedback back into the dataset.

```{r map Singer feedback, dev='png'}
fishData <- read.csv("data/processedFishDataWithDates.csv")
singerFeedback <- read.csv("missingFamilies_RandyReview.csv")

for (i in 1:nrow(singerFeedback)){
  update <- singerFeedback[i,]
  fishData[fishData$dwc_scientificName == update$was,1:3] <- update[1:3]
}

fishData <- fishData[complete.cases(fishData$dwc_family),]
fishData <- do.call(data.frame,fishData)
write.csv(fishData, 
          "data/processedFishDataWithDatesAndNames.csv", 
          row.names = F)
```

# Final results

Here is what was left at the end of processing, ready for analysis.

```{r}
paste0("There are ", nrow(fishData), " entries in the dataset.")
paste0("There are ", length(unique(fishData$dwc_scientificName)), 
       " species in the dataset.")
```
